---
title: "markov-inheritance"
author: "Amba Sharma"
date: "2026-06-03"
output: html_document
---

# Estimating an Inheritance with a Markov / Monte Carlo Model

## University Project: Markov Processes - reproducible analysis

Problem: a 60-year-old holds S0 = 1000 in a single FTSE-100 stock. The stock is sold on the day he dies and the proceeds are inherited. We estimate the distribution of the inherited sum S = S_N, where N is the (random) number of remaining years of life.

log S = log S0 + sum\_{i=1}\^{N} X_i, X_i = log(S_i / S\_{i-1})

#### Step 1: model N with a Gompertz distribution (fit to UK mortality data)

#### Step 2: model X_i as Normal (parameters via the method of percentiles)

#### Step 3: combine via Monte Carlo simulation

```{r}
set.seed(2024)                     # reproducibility
S0      <- 1000                    # initial investment
AGE0    <- 60                      # current age
N_SIMS  <- 100000                  # Monte Carlo iterations
```

### PART 1. Distribution of N (remaining lifetime) via the Gompertz model

Gompertz hazard (force of mortality) at duration t years past age

AGE0: mu(t) = b \* exp(eta \* t)

Survival function: S(t) = exp( -(b/eta) \* (exp(eta \* t) - 1) )

log(hazard) is linear in t, so estimate (b, eta) by regressing log(mortality rate) on duration.

```{r}
life  <- read.csv("data/uk_life_table.csv")           # age, mx, qx, lx, dx, ex
adult <- subset(life, age >= AGE0)
t      <- adult$age - AGE0                             # years past age 60
log_mu <- log(adult$mx)                                # log central mortality rate
 
fit  <- lm(log_mu ~ t)
eta  <- unname(coef(fit)[2])                           # shape  (slope)
b    <- unname(exp(coef(fit)[1]))                      # baseline hazard (intercept)
cat(sprintf("Gompertz fit:  b = %.6f   eta = %.5f\n", b, eta))

# Sanity check against published life expectancy at 60 (ex column)
gompertz_survival <- function(t) exp(-(b / eta) * (exp(eta * t) - 1))
ex60_model <- integrate(gompertz_survival, 0, 60)$value
cat(sprintf("Modelled e(60) = %.2f yrs  |  life-table e(60) = %.2f yrs\n",
            ex60_model, adult$ex[1]))
 
# Sample N by inverting the Gompertz CDF
sample_gompertz <- function(n) {
  u <- runif(n)
  t <- (1 / eta) * log(1 - (eta / b) * log(u))
  pmax(round(t), 0)                                    # whole years, >= 0
}
```

### PART 2. Yearly log-returns X_i for Tesco (Normal, method of percentiles)

Method of percentiles: match two sample quantiles to standard-Normal z-scores to back out (mu, sigma).

```{r}

estimate_normal_by_percentiles <- function(x, p_lo = 0.25, p_hi = 0.75) {
  q_lo <- as.numeric(quantile(x, p_lo))
  q_hi <- as.numeric(quantile(x, p_hi))
  z_lo <- qnorm(p_lo); z_hi <- qnorm(p_hi)
  sigma <- (q_hi - q_lo) / (z_hi - z_lo)
  mu    <- q_hi - z_hi * sigma
  c(mu = mu, sigma = sigma)
}
 
# Pull Tesco PLC prices directly from Yahoo Finance.
# Tesco trades on the London Stock Exchange under the ticker "TSCO.L".
library(quantmod)
 
tsco <- getSymbols("TSCO.L", src = "yahoo",
                   from = "1984-01-01", to = "2024-12-31",
                   auto.assign = FALSE)
 
# Collapse daily prices to one year-end (adjusted) close per year, so that
# X_i = log(S_i / S_{i-1}) is a yearly log-return, matching the brief.
yearly_price <- to.yearly(Ad(tsco), OHLC = FALSE)      # year-end adjusted close
log_returns  <- diff(log(as.numeric(yearly_price)))    # yearly log-returns
 
params <- estimate_normal_by_percentiles(log_returns)
mu     <- unname(params["mu"])
sigma  <- unname(params["sigma"])
cat(sprintf("Tesco yearly log-returns: n = %d,  mu = %.4f,  sigma = %.4f\n",
            length(log_returns), mu, sigma))
```

### PART 3. Monte Carlo for the distribution of log S

Each simulation: draw N, draw N i.i.d. X_i \~ Normal(mu, sigma), log S = log S0 + sum(X_i).

Repeat N_SIMS times.

```{r}
N_draws <- sample_gompertz(N_SIMS)
logS <- log(S0) + vapply(
  N_draws,
  function(n) if (n == 0) 0 else sum(rnorm(n, mu, sigma)),
  numeric(1)
)
 
mean_logS <- mean(logS); var_logS <- var(logS)
 
# Closed-form cross-check (law of total mean / variance):
#   E[log S]   = log S0 + E[N] * mu
#   Var[log S] = E[N] * sigma^2 + Var[N] * mu^2
EN <- mean(N_draws); VN <- var(N_draws)
mean_cf <- log(S0) + EN * mu
var_cf  <- EN * sigma^2 + VN * mu^2
```

### Results

```{r}
cat("\n Results \n")
cat(sprintf("E[N]               = %.2f years\n", EN))
cat(sprintf("E[log S]           = %.4f   (closed form %.4f)\n", mean_logS, mean_cf))
cat(sprintf("Var[log S]         = %.4f   (closed form %.4f)\n", var_logS, var_cf))
cat(sprintf("Median inherited S = %.2f\n", exp(median(logS))))
cat(sprintf("Mean   inherited S = %.2f\n", mean(exp(logS))))
qS <- quantile(exp(logS), c(0.05, 0.95))
cat(sprintf("90%% interval for S = [%.0f, %.0f]\n", qS[1], qS[2]))
```

### Figures

```{r}
dir.create("figures", showWarnings = FALSE)
 
png("figures/mortality_fit.png", width = 900, height = 600)
plot(adult$age, adult$mx, log = "y", pch = 16, col = "steelblue",
     xlab = "Age", ylab = "Central mortality rate (log scale)",
     main = "Gompertz fit to UK mortality, ages 60+")
lines(adult$age, b * exp(eta * t), col = "firebrick", lwd = 2)
legend("topleft", c("Observed", "Gompertz fit"),
       col = c("steelblue", "firebrick"), pch = c(16, NA), lwd = c(NA, 2))
dev.off()
 
png("figures/logS_distribution.png", width = 900, height = 600)
hist(logS, breaks = 60, col = "grey85", border = "white",
     xlab = "log S", main = "Monte Carlo distribution of log S")
abline(v = mean_logS, col = "firebrick", lwd = 2)
dev.off()
 
cat("\nFigures written to figures/.\n")
 
```
